---
title: "Getting started with bfo.db"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with bfo.db}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # The examples connect to a live database, so they are shown but not run
  # when the vignette is built.
  eval = FALSE
)
```

This guide walks you through everything you need to connect to the Barefoot
Ocean database and pull data into R. **No prior package-development experience
is assumed** — if you can run code in RStudio, you can follow along.

`{bfo.db}` gives you one object, `DatabaseConnector`, that:

* opens a secure (encrypted) connection to the database,
* lets you read database *views* into R, and
* cleans up the connection when you're done.

```{r setup}
library(bfo.db)
```

# 1. What you'll need

Before connecting, get these five values from the Barefoot Ocean data team:

| Field      | Meaning                | Example                                    |
|------------|------------------------|--------------------------------------------|
| `dbname`   | the database name      | `bfo`                                      |
| `host`     | the server address     | `bfo.abc123.us-east-1.rds.amazonaws.com`   |
| `port`     | the network port       | `5432`                                     |
| `user`     | your database username | `jane.doe`                                 |
| `password` | your database password | (secret)                                   |

You do **not** need to obtain an SSL certificate — the package already ships
one and uses it for you (see [Choosing a certificate](#choosing-a-certificate)).

> **The one rule that matters:** your `user` and `password` are secrets. Never
> put them in a file that is shared, emailed, or committed to GitHub. The
> recommended methods below keep them on your computer only.

# 2. Choosing how to store your credentials

There are three ways to provide credentials. **If you're unsure, use Method A.**

| Method                            | Best for                              |
|-----------------------------------|---------------------------------------|
| **A.** a `config.yml` file        | most people; one analyst, one machine |
| **B.** `config.yml` + `.Renviron` | teams / shared projects / many users  |
| **C.** typed into the script      | quick one-off tests only              |

## Method A — a `config.yml` file

A `config.yml` is a small settings file you create once.

**Step 1.** Create and open it (in RStudio):

```{r}
file.edit("config.yml")
```

**Step 2.** Paste this in and replace each `XXXXXX` with your real values. YAML
relies on indentation, so keep the two-space indents and don't use tabs:

```yaml
default:
  bfo_data:
    dbname: XXXXXX
    host: XXXXXX
    port: XXXXXX
    user: XXXXXX
    password: XXXXXX
```

Save the file.

**Step 3.** Connect:

```{r}
db <- DatabaseConnector$new(config_path = "config.yml")
```

**Step 4.** Keep it private. If your project uses Git, make sure the file is
never committed:

```{r}
usethis::use_git_ignore("config.yml")
```

## Method B — `config.yml` + `.Renviron` (for teams and many users)

When several people share one project, you don't want everyone's password in a
shared `config.yml`. Instead:

* keep a **single, shareable `config.yml`** that holds *no* secrets — it only
  *points to* environment variables; and
* let each person keep their own secrets in a personal **`.Renviron`** file.

The same project then works for **many users**, and nobody's password leaves
their own machine.

**Step 1.** Open your personal `.Renviron`. The `usethis` package makes this
easy — it finds and opens the file for you, wherever it lives:

```{r}
# install.packages("usethis")     # if needed
usethis::edit_r_environ()              # your personal (user-level) .Renviron
# usethis::edit_r_environ("project")   # or one scoped to the current project
```

`.Renviron` is **not** R code — it's just `NAME=value` lines, with no quotes and
no spaces around the `=`:

```
BFO_DB_NAME=bfo
BFO_DB_HOST=bfo.abc123.us-east-1.rds.amazonaws.com
BFO_DB_PORT=5432
BFO_DB_USER=jane.doe
BFO_DB_PASSWORD=your-secret-password
```

**Step 2.** Save the file and **restart R** (RStudio: *Session → Restart R*).
`.Renviron` is only read when R starts.

**Step 3.** Write a `config.yml` that *reads* those variables. The `!expr` tag
runs a small R expression; `Sys.getenv("…")` looks up a value from `.Renviron`:

```yaml
default:
  bfo_data:
    dbname: !expr Sys.getenv("BFO_DB_NAME")
    host: !expr Sys.getenv("BFO_DB_HOST")
    port: !expr Sys.getenv("BFO_DB_PORT")
    user: !expr Sys.getenv("BFO_DB_USER")
    password: !expr Sys.getenv("BFO_DB_PASSWORD")
```

This file has no real secrets in it, so it is safe to commit and share.

**Step 4.** Connect, exactly as in Method A:

```{r}
db <- DatabaseConnector$new(config_path = "config.yml")
```

### Several profiles in one file

`config.yml` can hold more than one named profile (say a read-only and an admin
login). Add extra top-level blocks and pick one with `R_CONFIG_ACTIVE`:

```yaml
default:
  bfo_data:
    dbname:   !expr Sys.getenv("BFO_DB_NAME")
    host:     !expr Sys.getenv("BFO_DB_HOST")
    port:     !expr Sys.getenv("BFO_DB_PORT")
    user:     !expr Sys.getenv("BFO_READONLY_USER")
    password: !expr Sys.getenv("BFO_READONLY_PASSWORD")

admin:
  bfo_data:
    dbname:   !expr Sys.getenv("BFO_DB_NAME")
    host:     !expr Sys.getenv("BFO_DB_HOST")
    port:     !expr Sys.getenv("BFO_DB_PORT")
    user:     !expr Sys.getenv("BFO_ADMIN_USER")
    password: !expr Sys.getenv("BFO_ADMIN_PASSWORD")
```

```{r}
Sys.setenv(R_CONFIG_ACTIVE = "admin")   # "default" is used when this is unset
db <- DatabaseConnector$new(config_path = "config.yml")
```

## Method C — type credentials directly (quick tests only)

Fastest to try, but the password ends up in your script — so only for
throwaway tests, and never save or share such a script.

```{r}
db <- DatabaseConnector$new(
  dbname   = "bfo",
  host     = "bfo.abc123.us-east-1.rds.amazonaws.com",
  port     = 5432,
  user     = "jane.doe",
  password = "your-secret-password",
  sslmode  = "require"
)
```

A safer variant reads the secrets from `.Renviron` so nothing is written down:

```{r}
db <- DatabaseConnector$new(
  dbname   = Sys.getenv("BFO_DB_NAME"),
  host     = Sys.getenv("BFO_DB_HOST"),
  port     = Sys.getenv("BFO_DB_PORT"),
  user     = Sys.getenv("BFO_DB_USER"),
  password = Sys.getenv("BFO_DB_PASSWORD")
)
```

# 3. Choosing a certificate {#choosing-a-certificate}

The connection is encrypted, which needs AWS's public Certificate Authority
bundle to verify the server. This is **not** a secret, and **you usually do
nothing** — the package ships the bundle and uses it automatically (the global
bundle, which covers all regions, falling back to `us-east-1`).

Only override it for special cases:

```{r}
# Force the region-specific bundle that ships with the package:
db <- DatabaseConnector$new(
  config_path = "config.yml"
)
# ...or, with direct parameters:
db <- DatabaseConnector$new(
  dbname = "bfo", host = "…", port = 5432, user = "…", password = "…",
  sslrootcert = system.file("cert", "us-east-1-bundle.pem", package = "bfo.db")
)

# Use your own certificate file:
db <- DatabaseConnector$new(
  dbname = "bfo", host = "…", port = 5432, user = "…", password = "…",
  sslrootcert = "/path/to/your-cert.pem"
)
```

In a `config.yml`, the equivalent setting is `ssl_key:` inside the `default:`
block (omit it to use the default).

# 4. Reading data

Once connected, you can work with a view in two ways.

**Lazily** (recommended): `get_view()` returns a reference — no data is fetched
until you `collect()`, so filters and joins run on the database, not in R:

```{r}
db$get_view("global_ref") |>
  dplyr::filter(country == "IDN") |>
  dplyr::collect()
```

**All at once**: `collect_view()` pulls the whole view into memory immediately:

```{r}
global_ref_data <- db$collect_view("global_ref")
```

Look up a single user:

```{r}
db$get_user_information(user_email = "jane.doe@barefootocean.org")
```

# 5. Closing the connection

When you're finished, close the pool:

```{r}
db$close()        # db$disconnect() is an alias
```

`close()` is safe to call more than once, and the connection is also closed
automatically if you forget — but closing explicitly is good practice.

# 6. Turning off the messages

By default the package prints a short explanation at each step. To silence
them, pass `quiet = TRUE` when connecting:

```{r}
db <- DatabaseConnector$new(config_path = "config.yml", quiet = TRUE)
```

# 7. Troubleshooting

| Message                                   | Usually means                         | Fix                                                   |
|-------------------------------------------|---------------------------------------|-------------------------------------------------------|
| `All connection parameters are required…` | a field is empty or misspelled        | check `config.yml`/`.Renviron`; restart R if you edited `.Renviron` |
| `could not connect to server` / timeout   | wrong `host`/`port`, or no VPN/network | confirm values and network access                    |
| `password authentication failed`          | wrong `user` or `password`            | re-check those two                                    |
| `.Renviron` values come back empty        | R wasn't restarted after editing it   | *Session → Restart R*                                 |

To confirm your `.Renviron` loaded (prints the host, not the password):

```{r}
Sys.getenv("BFO_DB_HOST")
```
