---
title: "Getting started with bfo.db"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with bfo.db}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # The examples connect to a live database, so they are shown but not run
  # when the vignette is built.
  eval = FALSE
)
```

`{bfo.db}` provides a small [R6](https://r6.r-lib.org/) interface,
`DatabaseConnector`, for connecting to the Barefoot Ocean PostgreSQL database
over a secure (TLS) connection and pulling database views into R.

```{r setup}
library(bfo.db)
```

## 1. Open a connection

There are two ways to connect. Either way, the package prints short messages
explaining what is happening and what to do next; pass `quiet = TRUE` to silence
them.

### Option A — a `config.yml` file (recommended)

Keep your secrets in a `config.yml` file that is **never committed to git**:

```yaml
default:
  bfo_data:
    dbname: XXXXXX
    host: XXXXXX
    user: XXXXXX
    password: XXXXXX
    port: XXXXXX
  # Optional: omit to use the CA bundle shipped with the package.
  # ssl_key: path/to/your-cert.pem
```

```{r connect-config}
db <- DatabaseConnector$new(config_path = "path/to/config.yml")
```

### Option B — direct parameters

```{r connect-direct}
db <- DatabaseConnector$new(
  dbname   = "XXXXXX",
  host     = "XXXXXX",
  user     = "XXXXXX",
  password = "XXXXXX",
  port     = 5432,
  sslmode  = "require"
  # sslrootcert is optional — defaults to the bundled AWS RDS CA bundle.
)
```

## 2. Credentials vs. the SSL certificate

Your database **host, user, and password are secrets** — keep them in
`config.yml` or `.Renviron` and out of version control.

The **SSL certificate** (`sslrootcert` / `ssl_key`) is *not* a secret: it is
AWS's public Certificate Authority bundle, used only to verify the server's
identity. For convenience this package ships the bundles itself, so by default
you don't need to download or manage any `.pem` file. The connection uses the
**global** bundle (which covers every AWS region), falling back to the
`us-east-1` bundle:

```{r cert-paths}
system.file("cert", "global-bundle.pem", package = "bfo.db")   # preferred
system.file("cert", "us-east-1-bundle.pem", package = "bfo.db") # fallback
```

Only set `sslrootcert`/`ssl_key` yourself to pin a specific certificate.

## 3. Fetch data

You can work with a view lazily (recommended) or pull it straight into memory.

### Lazily, then collect

`get_view()` returns a lazy reference — no data is fetched until you
`collect()`, so you can push filters and joins down to the database:

```{r get-view}
global_ref <- db$get_view("global_ref")

global_ref |>
  dplyr::filter(country == "IDN") |>
  dplyr::collect()
```

### All at once

```{r collect-view}
global_ref_data <- db$collect_view("global_ref")
```

### Look up a user

```{r user-info}
db$get_user_information(user_email = "anastasiia@barefootocean.org")
```

## 4. Close the connection

When you're done, return the pooled connections and close the pool:

```{r close}
db$close()        # db$disconnect() is an alias
```

`close()` is safe to call more than once, and the pool is also closed
automatically if the object is garbage-collected — but closing it explicitly is
good practice.
